SHARP II: Dark Matter Substructure, Gravitational Lens flux-ratio Anomalies, and the Case of B1555+375: Beware the Effect of Edge-on Disks

### 2015-05-03 papers/B1555_EdgeOn/B1555_letter.tex
update of modelling approach and Dave's comment

It was first suggested by \citet{Mao1998} that flux-ratio anomalies
observed in radio-loud multiple images of lensed quasars could be
interpreted as a telltale sign of the presence of substructure in lens
galaxies.  Indeed, small perturbations in the gravitational potential
are sufficient to cause flux-ratio anomalies in strong lensed systems
\citep{Bradac02}.  Moreover, unlike in the optical or infrared, in the
radio band, lensed images are not sensitive to microlensing from
stars nor dust extinction. This makes flux-ratio anomalies a promising
tool to detect \textbf{and measure} dark matter substructures \citep{Dalal2002}.  At
present, the amount of substructure derived from flux-ratio anomalies
is in disagreement with predictions from numerical simulations, with
the former requiring a larger fraction of mass in substructure than
expected \citep{Xu14}. While this could be the result of small sample
statistics, it could also be an indication that flux-ratio anomalies
have a different origin than substructure.
%
An alternative method for
the detection of substructure in gravitational lens galaxies based on
the surface brightness distribution of extended arcs and Einstein rings
has been introduced by \citet{K05} and \citet{V09}. Interestingly, the fraction of
substructure measured with this \emph{gravitational imaging technique}
is currently in agreement with theoretical expectations \citep{V14a,V12}.

@@

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot \textbf{reproduce} the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE). [WHY MODEL BULGE W/ SIE? COMMENT ON THE REASON FOR PICKING MODELS--LEON]
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is due to that some lens model parameters contain distance information. Our choice of redshifts does not affects the predicted magnifications nor flux-ratios.}

Our steps to explore the parameter space are: (1) Apply
strong priors from the SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge\textbf{/halo} component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The \textbf{final} best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modelling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

@@

Gravitational lensing is, at present, the most powerful tool for
studying substructure in galaxies outside of the Local Universe, and
the flux-ratio anomaly systems are a key aspect of these
investigations.  However, before using the flux-ratio anomaly in any
given system to investigate structure formation and the properties of
dark matter, it is critical to eliminate other possible explanations
for the difference between the observed and expected fluxes, such as
propagation effects and assumptions about the mass distribution of the
lens galaxy.  It is for this reason that much of the effort in the
flux-ratio technique has been focused on radio-loud lens systems.  At
these wavelengths, dust extinction and microlensing effects should be
minimal \citep[although see][for a rare example of radio
  microlensing]{K2000}.  \textbf{Also,  scatter broadening} can affect radio
fluxes, as is clearly shown in the B0128+437 system \citep{B04}, and free-free absorption can change the flux-ratios as a function of frequency,
for example, B0218+357 \citep{M07}.
However, these propagation effects can be disentangled from
substructure lensing because they have well known dependencies on
wavelength. \citet{KD04} examine all eight radio-loud flux anomalous
lenses besides B0128+437, namely: MG 0414+0534, B0712+472, B1359+154,
B1422+231, B1555+375, B1608+656, B1933+503, and B2045+265.  They found
no strong wavelength dependence in the flux-ratios of all these
systems and therefore rule out radio propagation effects as
significant contributors to the observed anomalies.

In this letter, we focus on the effect of more complicated mass
distributions for the lens galaxy on flux-ratio anomalies.
Traditionally, flux anomalous lensed quasars are radio-loud systems
that are firstly discovered in radio observations. While radio data
provide reliable information on the positions and the fluxes of the
lensed images, they are often unable to detect the lens
galaxies. Moreover, lens modelling based only on the positions of the
four images can be highly degenerate and prior assumptions play a key
role in the interpretation of the observed flux-ratio anomaly.  Since
lens galaxies are typically at redshifts of $z \leq 1$, optical and
near-infrared observations are ideal for measuring the light distribution
of lens galaxies and obtaining useful prior information on the underlying
mass distributions. Indeed \citet{SHARP12} showed that high
resolution IR imaging can help to significantly improve the precision
of the lens modelling. This improvement is even more dramatic in the
case of B1555+375, where \textit{HST} and SHARP AO high resolution imaging has
allowed us to detect, for the fist time, the presence of an edge-on
disk. This newly discovered edge-on disk provides a strong prior for our
lens modelling.  With \textbf{an additional disk component}, our models produce
matches to both the radio image positions and their flux-ratio, while
models with only one SIE component \citep{Marlow99,Xu14} fail to
reproduce the fluxes correctly. Our successful modelling indicates
that the kiloparsec-scale structure of the lens galaxy itself is
sufficient to explain the flux-ratio anomalies in B1555+375, without
any need to invoke the presence of substructure.

@@

Our results clearly
demonstrate that the lack of knowledge on the lens galaxy mass
distribution, combined with the limited information provided by the
position of a few lensed images, can lead to improper lens models and
a misinterpretation of the origin of flux-ratio anomalies. \textbf{While future large-scale surveys, such as the DES, KiDS-VIKING, Euclid, and LSST, are expected to
deliver new large samples of gravitationally lensed quasars, in the
near future the discovery of new systems using narrow-line quasar
emission \citep{N14} will provide new invaluable statistical
perspective on current constraints based on flux-ratio anomalies.} We stress,
therefore, that one should be very careful when interpreting the
nature of flux-ratio anomalies, as prior assumptions on the lens mass
distribution are a significant source of bias. Our analysis shows that
in order to derive precise quantification of the substructure
population and any derived properties from strongly lensed flux
anomalous systems, it is crucial to have a careful estimation of
non-substructure effects. In fact, even in the presence of mass
substructures, the observed flux-ratio anomalies could be still
dominated by the complex structure of the lens galaxy. Finally,
additional near-infrared AO imaging from SHARP shows that B1555+375 is
not the only system having a significant disk component detection in
the lensing galaxy.  These systems will be investigated in future
work, but the images alone strongly suggest that other lens systems
have similar links between flux-ratio anomalies and disk components.

>>>>

It was first suggested by \citet{Mao1998} that flux-ratio anomalies
observed in radio-loud multiple images of lensed quasars could be
interpreted as a telltale sign of the presence of substructure in lens
galaxies.  Indeed, small perturbations in the gravitational potential
are sufficient to cause flux-ratio anomalies in strong lensed systems
\citep{Bradac02}.  Moreover, unlike in the optical or infrared, in the
radio band, lensed images are neither sensitive to microlensing from
stars, nor to dust extinction. This makes flux-ratio anomalies a promising
tool to detect \textbf{and measure} dark matter substructures \citep{Dalal2002}.  At
present, the amount of substructure derived from flux-ratio anomalies
is in disagreement with predictions from numerical simulations, with
the former requiring a larger fraction of mass in substructure than
expected \citep{Xu14}. While this could be the result of small sample
statistics, it could also be an indication that flux-ratio anomalies
have a different origin than substructure.
%
An alternative method for
the detection of substructure in gravitational lens galaxies, utilizing
the surface brightness distribution of extended arcs and Einstein rings
has been introduced by \citet{K05} and \citet{V09}. Interestingly, the fraction of
substructure measured with this \emph{gravitational imaging technique}
is currently in agreement with theoretical expectations \citep{V14a,V12}.

@@

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot \textbf{reproduce} the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE). [WHY MODEL BULGE W/ SIE? COMMENT ON THE REASON FOR PICKING MODELS--LEON]
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is needed because some lens model parameters contain distance information. However, our choice of redshifts does not affect the predicted magnifications nor flux-ratios.}

\textbf{For each model, we explore the parameter space in the following way: 
(1) Preparation of priors: Applying the information in the SHARP AO image, we set ellipticity $e=0.8$ and position angle $\theta=10 \deg$ to the edge-on disk component. We then use the best-fit model of \citet{Marlow99} as the starting point for the bulge\textbf{/halo} component and also the source position. The rest of parameters in the disk component are set to be the same as the bulge/halo component as the starting point.
(2) Markov Chain Monte Carlo: In {\tt glafic}, we run MCMC to find the a proper set of parameters for further optimization. The smallest $\chi^2$ doesn't necessarily give a physical model. Here are the examples we would like to discard: the ellipticity and the position angle of the disk component are far from what we see in the SHARP AO image, the center positions of two components are too far apart physically, or the positions of the lenses and the source are clearly located outside the Einstein radius in the image.
(3) Optimize for physical lens models: We bind the positions of the two components with a strong Gaussian prior having the same central position and $\sigma = 0.05 $ arcsec and then run optimize [CHECK THE ALGORITHM?] to reach the final best-fit for each model.}

The \textbf{final} best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modelling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

@@

Gravitational lensing is, at present, the most powerful tool for
studying substructure in galaxies outside of the Local Universe, and
the flux-ratio anomaly systems are a key aspect of these
investigations.  However, before using the anomaly in any
given system to investigate structure formation and the properties of
dark matter, it is critical to eliminate other possible explanations
for the difference between the observed and expected fluxes, such as
propagation effects and assumptions about the lens galaxy's mass distribution.  
It is for this reason that much of the effort in the
flux-ratio technique has been focused on radio-loud lens systems.  At
these wavelengths, dust extinction and microlensing effects should be
minimal \citep[although see][for a rare example of radio
  microlensing]{K2000}.  \textbf{Also,  scatter broadening} can affect radio
fluxes, as is clearly shown in the B0128+437 system \citep{B04}, and free-free absorption can change the flux-ratios as a function of frequency,
for example, B0218+357 \citep{M07}.
However, these propagation effects can be disentangled from
substructure lensing because they have well known dependencies on
wavelength. In particular, \citet{KD04} examine all eight radio-loud flux anomalous
lenses besides B0128+437, namely: MG 0414+0534, B0712+472, B1359+154,
B1422+231, B1555+375, B1608+656, B1933+503, and B2045+265.  They find
no strong wavelength dependence in the flux-ratios of all these
systems and therefore rule out radio propagation effects as
significant contributors to the observed anomalies.

In this letter, we focus on the effect of more complicated mass
distributions for the lens galaxy on flux-ratio anomalies.
Traditionally, flux anomalous lensed quasars are radio-loud systems
that are firstly discovered in radio observations. While radio data
provide reliable information on the positions and the fluxes of the
lensed images, they are often unable to detect the lens
galaxies. Moreover, lens modelling based only on the positions of the
four images can be highly degenerate and prior assumptions play a key
role in the interpretation of the observed flux-ratio anomaly.  Since
lens galaxies are typically at redshifts of $z \leq 1$, optical and
near-infrared observations are ideal for measuring the light distribution
of lens galaxies and obtaining useful prior information on the underlying
mass distributions. Indeed \citet{SHARP12} showed that high
resolution IR imaging can help to significantly improve the precision
of the lens modelling. This improvement is even more dramatic in the
case of B1555+375, where \textit{HST} and SHARP AO high resolution imaging has
allowed us to detect, for the fist time, the presence of an edge-on
disk. This newly discovered edge-on disk provides a strong prior for our
lens modelling.  With \textbf{an additional disk component}, our models produce
matches to both the radio image positions and their flux-ratio, wheras
models with only one SIE component \citep{Marlow99,Xu14} fail to
reproduce the fluxes correctly. Our successful modelling indicates
that the kiloparsec-scale structure of the lens galaxy itself is
sufficient to explain the flux-ratio anomalies in B1555+375, without
any need to invoke the presence of substructure.

@@

Our results clearly
demonstrate that the lack of knowledge on the lens galaxy mass
distribution, combined with the limited information provided by the
position of a few lensed images, can lead to improper lens models and
a misinterpretation of the origin of flux-ratio anomalies. \textbf{While upcoming large-scale surveys, such as the DES, KiDS-VIKING, Euclid, and LSST, are expected to
deliver new large samples of gravitationally lensed quasars, in the
near future the discovery of new systems using narrow-line quasar
emission \citep{N14} will provide new invaluable statistical
perspective on current constraints based on flux-ratio anomalies.} We stress,
therefore, that one should be very careful when interpreting the
nature of flux-ratio anomalies, as prior assumptions on the lens mass
distribution are a significant source of bias. Our analysis shows that
in order to derive precise quantification of the substructure
population and any derived properties from strongly lensed flux
anomalous systems, it is crucial to have a careful estimation of
non-substructure effects. In fact, even in the presence of mass
substructures, the observed flux-ratio anomalies could be still
dominated by the complex structure of the lens galaxy. Finally,
additional near-infrared AO imaging from SHARP shows that B1555+375 is
not the only system having a significant disk component detection in
the lensing galaxy.  These systems will be investigated in future
work, but the images alone strongly suggest that other lens systems
have similar links between flux-ratio anomalies and disk components.
